---
title: "PCA"
author: "Ana"
date: "14/6/2020"
output:
  word_document: default
  html_document: default
---
1.PCA with FactoMineR function
```{r}
library(dplyr)
library (ggplot2)
library(FactoMineR)

final_data <- read.csv("C:/Users/Ana/Desktop/final_data.csv")
sapply(final_data, class)
#remove factor and responsive vble
fd_for_pca <- select(final_data, 
                     -countries_en, 
                     -nutriscore_grade,
                     -pnns_groups_2,
                     -nutriscore_score)

head(final_data, n=9)
#data for PCA 36556 - 8 vbles


#PCA con funcion
pca_final_data <- prcomp(fd_for_pca, center = TRUE, scale. = TRUE)
summary(pca_final_data)
biplot(pca_final_data, choices=1:2)
library(factoextra)
fviz_pca_var(pca_final_data,
             col.var = "contrib", 
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE     
             )
#compute standard deviation for each component
standard_deviation <- pca_final_data$sdev
#compute variance
var <- standard_deviation^2
#variance of 4 first PCA
var[1:6]
#proportion of variance explained
prop_varex <- var/sum(var)
prop_varex[1:8]
plot(prop_varex, xlab = "Principal Component",
             ylab = "Proportion of Variance Explained",
             type = "b")


library(corrplot)
var <- get_pca_var(pca_final_data)
corrplot(var$cos2, is.corr=FALSE)
fviz_cos2(pca_final_data, choice = "var", axes = 1:2)

```
according to last elbow rule, the first 4 PCA are significant, retaining a total of 73% of inertia

The most well represented feature is carbohydrates along sugas as these two are usually representative as the same nutrient in foods. 


the least represented dimension is salt, as the value registered in the dataset corresponds to a 100g ration of the product, and the retio of this ingredient in that amount of food is usually not representative.


2.PCA function with categorial dimensions as sumplementary
```{r}
#PCA con factors as supplementary to plot in same factorial plane
fd1 <- select(final_data,-nutriscore_score)
pca1 <- PCA(fd1,quali.sup=c(1,3,5)) 

fviz_pca_biplot(pca1, label = "var", habillage=fd1$nutriscore_grade)

#fviz_pca_biplot(pca1, label = "var", habillage=fd1$countries_en)

```
high nutrition score grades are mostly foods with high protein values


3. PCA buil function with weights
```{r}
#PCA con pesos 

library(magrittr)
library(ggpubr)

X <- as.matrix(fd_for_pca)
n <- nrow(X); varnames <- colnames(X)
p <- ncol(X); indnames <- rownames(X)

n <- nrow(X); varnames <- colnames(X)
p <- ncol(X); indnames <- rownames(X)

custom_PCA <- function(X, weights, metric="eucl", graph=FALSE) {
    N <- diag(x=weights*1/sum(weights), nrow=n, ncol=n)
    G <- apply(N %*% X, 2, function(var) sum(var))
    Xcent <- sweep(X,2,G)
    if (metric == "eucl") {
      cov <- t(Xcent) %*% N %*% Xcent
      eigen <- eigen(cov); X <- Xcent
    } else {
      s <- apply(X, 2, function(var) sd(var[weights == 1]))
      S <- diag(x=1/s, nrow=p, ncol=p)
      Xstan <- Xcent %*% S
      cor <- t(Xstan) %*% N %*% Xstan
      eigen <- eigen(cor); X <- Xstan
    }
    
    eigenvalues <- eigen$values
    eigenvectors <- eigen$vectors
    kaiser.rule.lim <- mean(eigenvalues)
    sign.comp <- which(eigenvalues > kaiser.rule.lim)
    ret.info <- sum(eigenvalues[sign.comp])/sum(eigenvalues)
    print(paste0(length(sign.comp)," significant components (", round(ret.info*100,digits=2), "%)"))
    proj.ind <- X %*% eigenvectors[,sign.comp]
    proj.var <- sqrt(eigenvalues)*eigenvectors[,sign.comp]
    
    ggplot_screeplot <- function(eigenvalues, kaiser.rule.lim) {
  p <- length(eigenvalues)
  ggplot() + geom_hline(yintercept=kaiser.rule.lim, col="red") +
    geom_point(aes(1:p,eigenvalues), col="blue", size=2.8)  +
    geom_line(aes(1:p,eigenvalues), col="blue", size=0.8) + customtheme(tsize=12) + 
    ggtitle("Scree Plot") + xlab("Component Number") + ylab("Eigenvalue")
}

ggplot_ind_fm <- function(eigenvalues, proj.ind, colind, indnames) {
  d1.expl <- eigenvalues[1]/sum(eigenvalues)*100
  d2.expl <- eigenvalues[2]/sum(eigenvalues)*100
  xlab <- xlab(paste0("Dim 1 (", round(d1.expl, digits=2), "%)"))
  ylab <- ylab(paste0("Dim 2 (", round(d2.expl, digits=2), "%)"))
  ggplot() + geom_factormap_theme(xintercept=0, yintercept=0) + customtheme() +
    geom_point(aes(x=proj.ind[,1], y=proj.ind[,2], col=factor(data$demo)), size=2) +
    geom_text(aes(x=proj.ind[,1], y=proj.ind[,2], label=indnames), hjust=0, vjust=-0.5) +
    ggtitle("Individuals factor map (PCA)") + labs(colour="Demo") + xlab + ylab
}

ggplot_var_fm <- function(eigenvalues, proj.var, varnames) {
  d1.expl <- eigenvalues[1]/sum(eigenvalues)*100
  d2.expl <- eigenvalues[2]/sum(eigenvalues)*100
  xlab <- xlab(paste0("Dim 1 (", round(d1.expl, digits=2), "%)"))
  ylab <- ylab(paste0("Dim 2 (", round(d2.expl, digits=2), "%)"))
  ggplot() + geom_factormap_theme(xintercept=0, yintercept=0, circle=T) + customtheme() +
    geom_segment(aes(x=0, y=0, xend=proj.var[,1], yend=proj.var[,2]), arrow=arrow(length=unit(0.3,"cm"))) +
    geom_text(aes(x=proj.var[,1], y=proj.var[,2], label=varnames), hjust=0.5, vjust=-0.5) +
    ggtitle("Variables factor map (PCA)") + xlab + ylab + coord_fixed(ratio=1)
}
customtheme <- function(tsize=12) {
    return (theme_bw() + 
              theme(panel.grid.major=element_blank(), panel.grid.minor=element_blank(),
                    axis.text.x=element_text(size=tsize), axis.text.y=element_text(size=tsize),
                    axis.title.x=element_text(size=tsize), axis.title.y=element_text(size=tsize),
                    plot.title=element_text(size=tsize+2, face="bold", hjust=0.5),
                    legend.text=element_text(size=tsize)) +
              theme(legend.position="bottom"))
  }
  
  geom_circle <- function(center=c(0,0), rad=1, npoints=100){
    c <- annotate("path",
                  x=c(0)+rad*cos(seq(0,2*pi,length.out=100)),
                  y=c(0)+rad*sin(seq(0,2*pi,length.out=100)))
    return(c)
  }
  
  geom_factormap_theme <- function(xintercept, yintercept, circle=FALSE) {
    hline <- geom_hline(yintercept=0, linetype="dashed")
    vline <- geom_vline(xintercept=0, linetype="dashed")
    if (circle) {
      circle <- geom_circle(center=c(0,0), rad=1, npoints=100)
      return(c(hline, vline, circle))
    } else {
      return(c(hline, vline))
    }
  }

    
    if (graph) {
        g1 <- ggplot_screeplot(eigenvalues, kaiser.rule.lim)
        g2 <- ggplot_ind_fm(eigenvalues, proj.ind, factor(data$demo), indnames)
        g3 <- ggplot_var_fm(eigenvalues, proj.var, varnames)
        print(ggarrange(g1,g2,g3,ncol=3,widths=c(2,4,2.5)))
    }
    return(list(proj.ind=proj.ind, proj.var=proj.var, eigenvalues=eigenvalues,
                eigenvectors=eigenvectors))
}

weights <- rep(1, n)
pca.all.eucl <- custom_PCA(X, weights=weights, metric="eucl", graph=FALSE)
pca.all.norm.eucl <- custom_PCA(X, weights=weights, metric="norm.eucl", graph=TRUE)
```

4. PCA with weights
```{r}
library(ggplot2)
sapply(fd1, class)
summary(pca1)
biplot(pca1)
pca1.eigenvalues <- pca1$eig[,c("eigenvalue")]
g1 <- ggplot_screeplot(pca1.eigenvalues, mean(pca1.eigenvalues))
g2 <- ggplot_ind_fm(pca1.eigenvalues, pca1$ind$coord, factor(fd1$demo), indnames)
g3 <- ggplot_var_fm(pca1.eigenvalues, pca1$var$coord, varnames)
ggarrange(g1,g2,g3,ncol=3,widths=c(2,4,2.5))
```

